$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-11-30 paper/paper.tex
Apply LIPIcs

\documentclass[sigplan,10pt,anonymous,review]{acmart}
\settopmatter{printfolios=true,printccs=false,printacmref=false}

@@

% Do not use italics in definitions and theorems
\theoremstyle{definition}
\newtheorem{nidefinition}{Definition}
\newtheorem{nitheorem}{Theorem}
\newtheorem{nilemma}{Lemma}
\newtheorem{example}{Example}
\renewcommand{\nidefinitionautorefname}{Definition}%
\renewcommand{\nitheoremautorefname}{Theorem}%
\renewcommand{\nilemmaautorefname}{Lemma}%
\renewcommand{\sectionautorefname}{\S}%
\renewcommand{\subsectionautorefname}{\S}%
\renewcommand{\exampleautorefname}{Example}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title[$\pi$ with leftovers: a mechanisation in Agda]{$\pi$ with leftovers: \\ a mechanisation in Agda}
\author{Uma Zalakain}
\affiliation{University of Glasgow, Scotland}
\email{u.zalakain.1@research.gla.ac.uk}
\orcid{https://orcid.org/0000-0002-3268-9338}{}
\author{Ornela Dardha}
\affiliation{University of Glasgow, Scotland}
\email{ornela.dardha@glasgow.ac.uk}
\orcid{https://orcid.org/0000-0001-9927-7875}{}
\keywords{pi-calculus, linear types, leftover typing, concurrency, mechanisation, Agda}
\setcopyright{acmlicensed}
\ccsdesc[300]{Theory of computation~Process calculi}
% \supplement{\url{https://github.com/umazalakain/typing-linear-pi}}
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10003752.10003753.10003761.10003764</concept_id>
<concept_desc>Theory of computation~Process calculi</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{acks}
We want to thank Erika Kreuter, Wen Kokke, James Wood, Guillaume Allais, Bob Atkey, and Conor McBride for their thoughts, patience, work, education and camaraderie.

The authors would also like to thank the anonymous referees fortheir valuable comments and helpful suggestions. This work is supported by the \grantsponsor{UKEPSRC}{UK EPSRC}{} grant \grantnum{UKEPSRC}{EP/K034413/1} ``From Data Types to Session Types: A Basis for Concurrency and Distribution'' (ABCD), and by the \grantsponsor{EUHORIZON}{EU HORIZON 2020 MSCA RISE}{} project \grantnum{EUHORIZON}{778233} ``Behavioural Application Program Interfaces'' (BehAPI).
\end{acks}

@@

  \item \textbf{Fully mechanised formalisation and type safety results}:
    The formalisation of the \picalc{} with leftover typing, from the syntax to the semantics and the type system, has been fully mechanised in Agda.

    We present the type safety results of our \picalc{} with leftovers in \autoref{type-safety}.
    We have fully mechanised \emph{framing} (\autoref{thm:framing}), \emph{weakening} (\autoref{thm:weakening}) and \emph{strengthening} (\autoref{thm:strengthening}).
    We use these results to then prove and mechanise \emph{subject congruence} (\autoref{thm:subject-congruence}), and together with \emph{substitution} (\autoref{thm:substitution}), to prove and mechanise \emph{subject reduction} (\autoref{thm:subject-reduction}).
\end{enumerate}
\begin{anonsuppress}
and is publicly available in our GitHub repository \cite{Zalakain2020Agda}
\end{anonsuppress}

@@

\paragraph{From names to de Bruijn indices}
When we translate into de Bruijn indices we keep the original binder names around --- they will serve as name hints for when we translate back.
The translation function $\func{fromRaw}$ works recursively, keeping a context $ctx : \Names_n$ that maps the first $n$ indices to their names.
Named references within the process are substituted with their corresponding de Bruijn index.
We demand that the original process is well-scoped: that all its free variable names appear in $ctx$ --- this is decidable and we therefore automate the construction of such a proof term.
\begin{alignat*}{2}
    &\func{fromRaw} && : (ctx : \Names_n) \; (P : \Raw) \\
    &               && \to \WellScoped \; ctx \; P \to \Process_n
\end{alignat*}

\paragraph{From de Bruijn indices to names}
The translation function $\func{toRaw}$ works recursively, keeping a context $ctx : \Names_n$ that maps the first $n$ indices to their names.
As some widely-used languages do, this translation function produces unique variable names.
These unique variable names use the naming scheme $<name-hint>^{<n>}$, where $^{<n>}$ denotes that the name $<name-hint>$ has already been bound $n$ times before.
\begin{equation*}
    \func{toRaw} : (ctx : \Names_n) \to \Process_n \to \Raw
\end{equation*}

@@

\begin{proof}[Proof]
  All the above results are proved by induction on \textsc{Process} and \textsc{Var} (\autoref{fig:syntax}).
  Complete details can be found in our mechanisation in Agda%
\begin{anonsuppress}
in \cite{Zalakain2020Agda}
\end{anonsuppress}
.
\end{proof}

@@

\begin{proof}[Proof]
  All the above theorems are proved by induction on \textsc{Types} and \textsc{VarRef}.
  For details, refer to our mechanisation in Agda%
\begin{anonsuppress}
\cite{Zalakain2020Agda}
\end{anonsuppress}
.
\end{proof}

@@

\begin{proof}[Proof]
  The proof is by induction on \textsc{Equals} $\eq{}$.
  Here we only consider those cases that are not purely inductive: the base cases for $\constr{struct}$ and their symmetric variants.
\begin{anonsuppress}
Full proof in \cite{Zalakain2020Agda}.
\end{anonsuppress}
  We proceed by induction on \textsc{StructCong} $\eqeq$:
  \begin{itemize}
    \item
      Case $\constr{comp-assoc}$: trivial, as leftover typing is naturally associative.
    \item
      Case $\constr{comp-sym}$ for $\comp{P}{Q}$: we use framing (\autoref{thm:framing}) to shift the output context of $P$ to the one of $Q$; and the input context of $Q$ to the one of $P$.
    \item
      Case $\constr{comp-end}$: trivial, as the typing rule for $\PO$ has the same input and output contexts.
    \item
      Case $\constr{scope-end}$: we show that the usage annotation of the newly created channel must be $\lz$, making the proof trivial.
      In the opposite direction, we instantiate the newly created channel to a type $\unit$ and a usage annotation $\lz$.
    \item
      Case $\constr{scope-ext}$ for $\new \comp{P}{Q}$: we to show that $P$ preserves the usage annotations of the unused variable (\autoref{lm:types-unused}) and then use strengthening (\autoref{thm:strengthening}).
      In the reverse direction, we use weakening (\autoref{thm:weakening}) on $P$ and show that lowering and then lifting $P$ results in $P$ (\autoref{lm:lower-lift}).
    \item
      Case $\constr{scope-comm}$:
      we use exchange (\autoref{thm:exchange}), and for the reverse direction exchange and \autoref{lm:exchange-exchange} to show that exchanging two elements in $P$ twice leaves $P$ unchanged. \qedhere
  \end{itemize}
\end{proof}

@@

\paragraph{Extrinsic Encodings}

@@

\paragraph{Intrinsic Encodings}

@@

\paragraph{Other Work}

@@

In this paper we provide a well-scoped syntax and a semantics for the \picalc{}, extrinsically define a type system on top of the syntax capable of handling linear, graded and shared types under the same unified framework and show that reduction preserves the well-typedness of processes.
We avoid the need for extrinsic context splits by defining a type system based on leftover typing \cite{Allais2018a}, which is defined here for the first time for the \picalc{}.
As a result, theorems like framing, weakening and strengthening can now we stated for the linear \picalc{}.
Our work is fully mechanised in around 1850 lines of code in Agda%
\begin{anonsuppress}
\cite{Zalakain2020Agda}
\end{anonsuppress}
.

>>>>

\documentclass[]{lipics-v2019}

@@

% Do not use italics in definitions and theorems
\theoremstyle{definition}
\newtheorem{nidefinition}{Definition}
\newtheorem{nitheorem}{Theorem}
\newtheorem{nilemma}{Lemma}
% \newtheorem{example}{Example}
\renewcommand{\nidefinitionautorefname}{Definition}%
\renewcommand{\nitheoremautorefname}{Theorem}%
\renewcommand{\nilemmaautorefname}{Lemma}%
\renewcommand{\sectionautorefname}{\S}%
\renewcommand{\subsectionautorefname}{\S}%
% \renewcommand{\exampleautorefname}{Example}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{$\pi$ with leftovers: \\ a mechanisation in Agda}
\titlerunning{$\pi$ with leftovers: a mechanisation in Agda}
\author{Uma Zalakain}{University of Glasgow, Scotland}{u.zalakain.1@research.gla.ac.uk}{https://orcid.org/0000-0002-3268-9338}{}
\author{Ornela Dardha}{University of Glasgow, Scotland}{ornela.dardha@glasgow.ac.uk}{https://orcid.org/0000-0001-9927-7875}{}
\keywords{pi-calculus, linear types, leftover typing, concurrency, mechanisation, Agda}
\Copyright{U. Zalakain, and O. Dardha}
\ccsdesc[300]{Theory of computation~Process calculi}
% \supplement{\url{https://github.com/umazalakain/typing-linear-pi}}
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10003752.10003753.10003761.10003764</concept_id>
<concept_desc>Theory of computation~Process calculi</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\acknowledgements{We want to thank Erika Kreuter, Wen Kokke, James Wood, Guillaume Allais, Bob Atkey, and Conor McBride for their thoughts, patience, work, education and camaraderie.}

\funding{The authors would also like to thank the anonymous referees for their valuable comments and helpful suggestions. This work is supported by the UK EPSRC grant EP/K034413/1 ``From Data Types to Session Types: A Basis for Concurrency and Distribution'' (ABCD), and by the EU HORIZON 2020 MSCA RISE project 778233 ``Behavioural Application Program Interfaces'' (BehAPI).}

@@

  \item \textbf{Fully mechanised formalisation and type safety results}:
    The formalisation of the \picalc{} with leftover typing, from the syntax to the semantics and the type system, has been fully mechanised in Agda and is publicly available in our GitHub repository \cite{Zalakain2020Agda}

    We present the type safety results of our \picalc{} with leftovers in \autoref{type-safety}.
    We have fully mechanised \emph{framing} (\autoref{thm:framing}), \emph{weakening} (\autoref{thm:weakening}) and \emph{strengthening} (\autoref{thm:strengthening}).
    We use these results to then prove and mechanise \emph{subject congruence} (\autoref{thm:subject-congruence}), and together with \emph{substitution} (\autoref{thm:substitution}), to prove and mechanise \emph{subject reduction} (\autoref{thm:subject-reduction}).
\end{enumerate}

@@

\paragraph*{From names to de Bruijn indices}
When we translate into de Bruijn indices we keep the original binder names around --- they will serve as name hints for when we translate back.
The translation function $\func{fromRaw}$ works recursively, keeping a context $ctx : \Names_n$ that maps the first $n$ indices to their names.
Named references within the process are substituted with their corresponding de Bruijn index.
We demand that the original process is well-scoped: that all its free variable names appear in $ctx$ --- this is decidable and we therefore automate the construction of such a proof term.
\begin{alignat*}{2}
    &\func{fromRaw} && : (ctx : \Names_n) \; (P : \Raw) \\
    &               && \to \WellScoped \; ctx \; P \to \Process_n
\end{alignat*}

\paragraph*{From de Bruijn indices to names}
The translation function $\func{toRaw}$ works recursively, keeping a context $ctx : \Names_n$ that maps the first $n$ indices to their names.
As some widely-used languages do, this translation function produces unique variable names.
These unique variable names use the naming scheme $<name-hint>^{<n>}$, where $^{<n>}$ denotes that the name $<name-hint>$ has already been bound $n$ times before.
\begin{equation*}
    \func{toRaw} : (ctx : \Names_n) \to \Process_n \to \Raw
\end{equation*}

@@

\begin{proof}[Proof]
  All the above results are proved by induction on \textsc{Process} and \textsc{Var} (\autoref{fig:syntax}).
  Complete details can be found in our mechanisation in Agda in \cite{Zalakain2020Agda}.
\end{proof}

@@

\begin{proof}[Proof]
  All the above theorems are proved by induction on \textsc{Types} and \textsc{VarRef}.
  For details, refer to our mechanisation in Agda \cite{Zalakain2020Agda}.
\end{proof}

@@

\begin{proof}[Proof]
  The proof is by induction on \textsc{Equals} $\eq{}$.
  Here we only consider those cases that are not purely inductive: the base cases for $\constr{struct}$ and their symmetric variants.
  Full proof in \cite{Zalakain2020Agda}.
  We proceed by induction on \textsc{StructCong} $\eqeq$:
  \begin{itemize}
    \item
      Case $\constr{comp-assoc}$: trivial, as leftover typing is naturally associative.
    \item
      Case $\constr{comp-sym}$ for $\comp{P}{Q}$: we use framing (\autoref{thm:framing}) to shift the output context of $P$ to the one of $Q$; and the input context of $Q$ to the one of $P$.
    \item
      Case $\constr{comp-end}$: trivial, as the typing rule for $\PO$ has the same input and output contexts.
    \item
      Case $\constr{scope-end}$: we show that the usage annotation of the newly created channel must be $\lz$, making the proof trivial.
      In the opposite direction, we instantiate the newly created channel to a type $\unit$ and a usage annotation $\lz$.
    \item
      Case $\constr{scope-ext}$ for $\new \comp{P}{Q}$: we to show that $P$ preserves the usage annotations of the unused variable (\autoref{lm:types-unused}) and then use strengthening (\autoref{thm:strengthening}).
      In the reverse direction, we use weakening (\autoref{thm:weakening}) on $P$ and show that lowering and then lifting $P$ results in $P$ (\autoref{lm:lower-lift}).
    \item
      Case $\constr{scope-comm}$:
      we use exchange (\autoref{thm:exchange}), and for the reverse direction exchange and \autoref{lm:exchange-exchange} to show that exchanging two elements in $P$ twice leaves $P$ unchanged. \qedhere
  \end{itemize}
\end{proof}

@@

\paragraph*{Extrinsic Encodings}

@@

\paragraph*{Intrinsic Encodings}

@@

\paragraph*{Other Work}

@@

In this paper we provide a well-scoped syntax and a semantics for the \picalc{}, extrinsically define a type system on top of the syntax capable of handling linear, graded and shared types under the same unified framework and show that reduction preserves the well-typedness of processes.
We avoid the need for extrinsic context splits by defining a type system based on leftover typing \cite{Allais2018a}, which is defined here for the first time for the \picalc{}.
As a result, theorems like framing, weakening and strengthening can now we stated for the linear \picalc{}.
Our work is fully mechanised in around 1850 lines of code in Agda \cite{Zalakain2020Agda}.
